---
title: "Real Estate Valuation: Regression Modeling and Diagnostics"
output: html_document
---

## Purpose

This analysis demonstrates how to build, simplify, diagnose, and evaluate a multiple-regression model for `PricePerSQFT`.

The source coursework identified a possible quadratic relationship for `HouseAge`, compared a full and reduced model, tested regression assumptions, evaluated multicollinearity and outliers, and assessed out-of-sample RMSE.

```{r setup, message=FALSE, warning=FALSE}
library(readxl)
library(caret)
library(ggplot2)
library(lmtest)
library(car)
library(leaps)

# Put RealEstateValue.xlsx in data/ before running.
RealEstateValue <- read_excel("data/RealEstateValue.xlsx")

set.seed(123)
training_samples <- createDataPartition(
  RealEstateValue$PricePerSQFT,
  p = 0.70,
  list = FALSE
)
train_data <- RealEstateValue[training_samples, ]
test_data  <- RealEstateValue[-training_samples, ]
```

## Exploratory relationships

```{r}
independent_vars <- setdiff(names(train_data), "PricePerSQFT")

for (var in independent_vars) {
  print(
    ggplot(train_data, aes(x = .data[[var]], y = PricePerSQFT)) +
      geom_point(alpha = 0.6) +
      geom_smooth(method = "lm", se = FALSE) +
      labs(
        title = paste("PricePerSQFT vs", var),
        x = var,
        y = "PricePerSQFT"
      ) +
      theme_minimal()
  )
}
```

The original analysis identified `HouseAge` as having a possible second-degree relationship with `PricePerSQFT`.

## Correlation screening

```{r}
predictors <- train_data[, setdiff(names(train_data), "PricePerSQFT")]
cor(predictors)
```

In the saved coursework results, the largest reported predictor correlations included:

- `ToTransitStation` and `Longitude`: approximately -0.79
- `ToTransitStation` and `NumConvenienceStores`: approximately -0.59

These correlations are screening signals, not by themselves reasons to automatically remove variables.

## Full regression model

```{r}
train_data$HouseAge2 <- train_data$HouseAge^2

full_model <- lm(
  PricePerSQFT ~ HouseAge + HouseAge2 + ToTransitStation +
    NumConvenienceStores + Latitude + Longitude,
  data = train_data
)

summary(full_model)

training_rmse <- function(model) {
  sqrt(mean(residuals(model)^2))
}

training_rmse(full_model)
AIC(full_model)
```

## Reduced model

The original backward-elimination step removed `Longitude`.

```{r}
reduced_model <- lm(
  PricePerSQFT ~ HouseAge + HouseAge2 + ToTransitStation +
    NumConvenienceStores + Latitude,
  data = train_data
)

summary(reduced_model)
training_rmse(reduced_model)
anova(reduced_model, full_model)
AIC(full_model, reduced_model)
```

The saved coursework reported approximately:

| Model | Adjusted R² | Training RMSE | Nested-model ANOVA | AIC |
|---|---:|---:|---:|---:|
| Full | 0.5863 | 76.91 | — | ~3371.7 |
| Reduced (-Longitude) | 0.5877 | 75.99 | p = 0.8104 | ~3371.7 |

These are preserved as **coursework-reported results** rather than newly recomputed metrics because the original Excel dataset is not included in this repository.

## Regression diagnostics

```{r}
par(mfrow = c(2, 2))
plot(reduced_model)
par(mfrow = c(1, 1))

qqnorm(residuals(reduced_model))
qqline(residuals(reduced_model))

shapiro.test(residuals(reduced_model))

dwtest(reduced_model)
acf(residuals(reduced_model), main = "ACF Plot of Residuals")

plot(
  fitted(reduced_model),
  residuals(reduced_model),
  xlab = "Fitted Values",
  ylab = "Residuals",
  main = "Residuals vs Fitted"
)
abline(h = 0)

crPlots(reduced_model)

ncvTest(reduced_model)
spreadLevelPlot(reduced_model)

vif(reduced_model)

outlierTest(reduced_model)
```

The original analysis reported:

- mild tail deviations in the Q-Q plot;
- no clear evidence of autocorrelated residuals;
- no obvious nonlinear residual pattern;
- evidence of heteroscedasticity (`ncvTest` reported p ≈ 0.003);
- high VIF values for `HouseAge` and `HouseAge2`;
- an outlier flagged at observation 193.

These findings matter because model diagnostics can reveal limitations even when headline fit statistics look acceptable.

## Polynomial-term multicollinearity

Including `HouseAge` and `HouseAge²` can naturally produce high VIF values because a variable and its square are mathematically related.

A common refinement is to center `HouseAge` before creating the quadratic term:

```{r}
train_data$HouseAge_centered <- train_data$HouseAge - mean(train_data$HouseAge)
train_data$HouseAge_centered2 <- train_data$HouseAge_centered^2

centered_model <- lm(
  PricePerSQFT ~ HouseAge_centered + HouseAge_centered2 +
    ToTransitStation + NumConvenienceStores + Latitude,
  data = train_data
)

vif(centered_model)
```

This is shown as a methodological extension; it was not part of the original exam model.

## All-subsets regression

```{r}
subset_model <- regsubsets(
  PricePerSQFT ~ HouseAge + HouseAge2 + ToTransitStation +
    NumConvenienceStores + Latitude,
  data = train_data,
  nbest = 1
)

subset_summary <- summary(subset_model)
subset_summary$adjr2
```

The saved coursework identified:

`PricePerSQFT ~ HouseAge + HouseAge2 + ToTransitStation + Latitude`

as the formula selected by adjusted R-squared in the all-subsets step.

## Held-out test evaluation

```{r}
test_data$HouseAge2 <- test_data$HouseAge^2
test_predictions <- predict(reduced_model, newdata = test_data)

test_rmse <- sqrt(
  mean((test_data$PricePerSQFT - test_predictions)^2)
)

test_rmse
```

The test RMSE should be treated as more relevant to predictive generalization than training residual RMSE because it evaluates observations not used to fit the model.

## Model-selection interpretation

The lowest training RMSE is **not automatically sufficient** to identify the best model.

A defensible choice should consider:

- held-out predictive performance;
- model complexity;
- statistical significance and nested-model comparisons;
- AIC or another appropriate information criterion;
- regression assumptions;
- multicollinearity;
- influential observations;
- interpretability and business purpose.

## Limitations

The source dataset is not included in the repository, so the numerical results preserved above are those recorded in the original coursework. A fully reproducible public portfolio version would require either permission to distribute the original dataset or a clearly documented public/synthetic substitute.
